\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-03-04 paper/main.tex
Pass through the text

\usepackage{url}
\usepackage{graphicx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{xspace}
\usepackage{pdflscape} 
\usepackage{multirow} 
\usepackage{multicol}
\usepackage{listings}
\usepackage{verbatim}

% Custom colors
\usepackage{color}
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

\definecolor{xlinkcolor}{rgb}{0.7752941176470588, 0.22078431372549023, 0.2262745098039215}
\definecolor{codegreen}{rgb}{0,0.4,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

@@

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pretraining), a method which associates astronomical observations imaged by astronomical surveys like the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, optionally summarized via guided generation using large language models (LLMs). We show that the model embodies a meaningful joint representation between observations and text through experiments targeting image retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

\SM{Add some qualitative evals with the abstract model.}

Machine learning (ML) is beginning to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on replacing or augmenting traditional techniques with ML counterparts in order to improve performance on specific, individual tasks.
%
The \emph{Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to tackle a wide range of tasks.
%
The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread use and adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP.
%
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains.
%
At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute.
%
However, foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it.
%
The multi-modality inherent to the nature of astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%
\textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

In this paper we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pretraining\footnote{Technically we fine tune rather than pretrain, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and corresponding downstream image observations. 
%
Concretely, here we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP (Contrastive Language-Image Pre-training)~\cite{radford2021learning} image-text model on observation-abstract associations results in meaningful joint representations through quantitative and qualitative evaluation.
%
This opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm.
%

The CLIP family of foundation models, which in their original instantiation embed images and associated captions into a common embedding space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval.
%
The concept of associating diverse modalities via contrastive training has been used on other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations.

@@

The paper is organized as follows.
%
In Sec.~\ref{sec:dataset}, we describe the dataset used in this work, including its curation and processing.
%
In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model.
%
In Sec.~\ref{sec:results}, we present the results of our experiments on image and text retrieval tasks.
%
We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions.
%
We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations, optionally summarizing the abstract via guided LLM generation (described in Sec.~\ref{sec:summarization} below).
%
The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.
%

Examples of images and corresponding (summarized) captions are shown in Tab.~\ref{tab:dataset}. % \SM{Push table earlier.}
%
It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking and processing which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models.
%
This further motivates the need for fine-tuning on domain-specific data.

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
      \toprule
      \centering \bfseries \hubble Image & \centering \bfseries Obs. cycle \\ (Year) & \centering \bfseries Prop. ID & \centering \bfseries Proposal abstract (clipped) \tabularnewline
      \midrule
      \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} \\ (2019) & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/abs1_0.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} \\ (2013) & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/abs1_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} \\ (1999) & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/abs1_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} \\ (2016) & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/abs1_3.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Examples of images and corresponding captions, constructed using the LLM-extracted summaries.
%
The CLIP model is fine-tuned using these text-image associations.
%
Although the summarized abstract is extracted as two parts (objects/phenomena and science use cases), the model is trained using the entire summary as a single caption.}
  \label{tab:dataset}
\end{table}

@@

We note that some fraction of the image-caption pairs in the constructed dataset will primarily concern instrumentation and/or calibration rather than scientific content.
%
We choose to not filter out these pairs from our dataset, in order to have a larger sample of HST observations that the image encoder can leverage to adapting to the distinctive characteristics of \hubble images.

@@

Raw proposal abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations.
%
These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well.
%
Although the abstracts can be used as-is as image captions, we experiment with summarizing them via guided large language model (LLM) generation to standardize the captions used for fine-tuning the CLIP model.
%
The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the autoregressive generation process by masking out tokens that would violate the intended format.
%
By framing the problem in terms of transitions between a set of finite-state machines, \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation.
%
This ensures that the output of the LLM strictly conforms to the format of the following example:

@@

which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation process.

We use the open-weights, instruction-tuned model \textsc{Mixtral-8x7B-Instruct}~\citep{jiang2024mixtral} to generate the summaries, with guided generation performed using the \package{Outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package.
%
While the schema-guided generation ensures the format of the output, the prompt and choice of LLM will influence the content of the generated summaries.
%
Further details on the summarization procedure, including the prompts and schema used, are provided in App.~\ref{app:summarization}.

We emphasize that the goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}.
%
We also note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images, as can be seen from Tab.~\ref{tab:dataset}.

@@

\begin{landscape}
      \begin{table}[h!]
        \renewcommand{\arraystretch}{2}
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the initial parts of raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (rightmost two columns), separately extracting objects and phenomena as well as potential downstream science use cases.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically  meaningful joint representation between HST observation images and natural language.
%
We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

\subsection{Contrastive Language-Image Pretraining (CLIP)}

Contrastive Language-Image Pretraining \citep[CLIP;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduce for training CLIP-style architectures \citep{radford2021learning}
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16 \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model has 149,620,737 trainable parameters.
%
The model was originally trained on 400 million image-text pairs from internet data.
%

\subsection{Fine-tuning procedure}

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data, the raw proposal abstracts, or single-concept summaries.
%
When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder.
%
Images are randomly cropped to the native resolution of the image encoder (maintaining $\sim 20\%$ of the area of the original image) and randomly rotated by discrete increments of $90^\circ$ at each training step.
%
Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were seen to be susceptible to overfitting.
%
We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract. The temperature hyperparameter $\tau$ was initialized to its pre-trained value.

We explore three different methods of training the model on our domain dataset: \emph{(1)} Fine-tuning the entire network, starting from the pre-trained base model; \emph{(2)} Freezing the base image and text encoders, and training a small projection head; and \emph{(3)} Training the entire model from scratch.
%
For \emph{(2)}, we use a 2-layer MLP with 1024 units and a GELU activation layer, projecting onto the 512-dimensional embedding space.

Models were trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate of $10^{-5}$ and weight decay $10^{-3}$.
%
Training takes approximately 3 hours on 4 Nvidia A100 GPUs.
Models were instantiated and trained using packages from the \package{Jax} \citep{jax2018github} ecosystem.
%

\subsection{Evaluation metrics}

The model is evaluated by tracking the loss in Eq.~\ref{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training.
%
The retrieval accuracy is defined as the fraction of associated captions which fall within the top $k\%$ of captions by cosine similarity of the normalized embeddings $x_i \cdot y_j$, averaged over the images in the validation set.

>>>>

\usepackage{url}
\usepackage{graphicx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{xspace}
\usepackage{pdflscape} 
\usepackage{multirow} 
\usepackage{multicol}
\usepackage{listings}
\usepackage{verbatim}
\usepackage{caption}
\usepackage{bbm}

% Custom colors
\usepackage{color}

\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}
\definecolor{deeppurple}{rgb}{0.60,0.42,0.63}

\definecolor{xlinkcolor}{rgb}{0.7752941176470588, 0.22078431372549023, 0.2262745098039215}

\definecolor{codegreen}{rgb}{0,0.4,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

@@

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training), a method which associates astronomical observations imaged by surveys and telescopes with natural language. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts, optionally summarized via guided generation using large language models (LLMs), and corresponding downstream observations. Using observations from the \hubble Space Telescope (HST) as an example, we show that the model embodies a meaningful joint representation between observations and text through experiments targeting image retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for finding patterns in astronomical data, in particular by leveraging text as an interface.
\end{abstract}

@@

% \SM{Add some qualitative evals with the abstract model.}

Machine learning (ML) is starting to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on augmenting traditional techniques in order to improve performance on specific tasks.
%
The {Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to simultaneously tackle a wide range of tasks \citep{bommasani2021opportunities}.
%
% The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread use and adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP.
%
% At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute.
% %
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised or weakly-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains.
%
However, foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it.
%
The multi-modality inherent to astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given target object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%
\textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

In this paper we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training\footnote{Technically we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and images corresponding to downstream observations. 
%
Concretely, here we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract associations results in meaningful joint representations through quantitative and qualitative evaluation.
%
This opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm.
%

The CLIP family of foundation models, which in their original form embed images and associated captions into a common representation space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval.
%
The concept of associating diverse modalities via contrastive training has been employed in other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations. Here we present for the first an application associating astronomical data with the text modality.

@@

The paper is organized as follows.
%
In Sec.~\ref{sec:dataset}, we describe the \hubble dataset used in this work, including the curation and processing of observations as well as text captions.
%
In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model.
%
In Sec.~\ref{sec:results}, we present quantitative and qualitative results of our experiments on image and text retrieval tasks.
%
We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions.
%
We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations, optionally summarizing the abstract via guided generation using LLMs (described in Sec.~\ref{sec:summarization} below).
%
The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.
%

Examples of images and corresponding (clipped) proposal abstracts are shown in Tab.~\ref{tab:dataset}. % \SM{Push table earlier.}
%
It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking and processing which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models \citep{deng2009imagenet}.
%
This further motivates the need for fine-tuning on domain-specific data.

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
      \toprule
      \centering \bfseries \hubble image & \centering \bfseries Obs. cycle \\ (Year) & \centering \bfseries Prop. ID & \centering \bfseries Proposal abstract (clipped) \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} \\ (2019) & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/abs1_0.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} \\ (2013) & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/abs1_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} \\ (1999) & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/abs1_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} \\ (2016) & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/abs1_3.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Examples of images (left-most column) and corresponding clipped proposal abstracts (right-most column). The observation cycle and corresponding year, as well as proposal ID are shown in the second and third columns, respectively. The proposal ID links to the Mikulski Archive for Space Telescopes (MAST) page corresponding to the proposal.}
  \label{tab:dataset}
\end{table}

@@

We note that some fraction of the image-caption pairs in the constructed dataset will primarily concern instrumentation and/or calibration rather than scientific content.
%
We choose to not filter out these pairs from our dataset, in order to have a larger sample of HST observations that the image encoder can leverage to adapt to the distinctive characteristics of \hubble images.

@@

Raw proposal abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations.
%
These abstracts are written in a diversity of styles, formats, and lengths while being highly variable in the nature of content.
%
Although the abstracts can be used as-is as image captions, we experiment with summarizing them via guided large language model (LLM) generation to standardize the captions used for fine-tuning the CLIP model.
%
The intended goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the autoregressive generation process by masking out tokens that would violate the intended format.
%
By framing the problem in terms of transitions between a set of finite states (a finite-state machine), \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation.
%
This ensures that, in this case, the output of the LLM strictly conforms to the format of the following example:

@@

which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation method used here.

While the schema-guided generation ensures the \emph{format} of the output, the prompt and choice of LLM will dictate the \emph{content} of the generated summaries.
%
We use the open-weights, instruction-tuned model \textsc{Mixtral-8x7B-Instruct}~\citep{jiang2024mixtral} to generate the summaries, with guided generation performed using the \package{Outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package.
%
Further details on the summarization procedure, including the prompts and schema used, are provided in App.~\ref{app:summarization}.

We emphasize that the intended goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}.
%
We also note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images, as can be seen from Tab.~\ref{tab:datasetsumm}.

@@

\begin{landscape}
      \begin{table}[h!]
        \renewcommand{\arraystretch}{2}
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6.5cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              % \midrule
              % \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the clipped raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (right-most two columns), separately extracting objects and phenomena as well as potential downstream science use cases.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically meaningful joint representation between images corresponding to HST observation and natural (English) language.
%
We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

\subsection{Contrastive Language-Image Pre-training (CLIP)}

CLIP \citep[Contrastive Language-Image Pre-training;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
The standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16\footnote{\url{https://huggingface.co/openai/clip-vit-base-patch16}} \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$ pixels.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model has 149,620,737 trainable parameters.
%
The model was originally trained on 400 million image-text pairs from internet data \citep{radford2021learning}.
%

\subsection{Fine-tuning Procedure}

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized abstracts or raw proposal abstracts.
%
When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder.
%
Images are augmented via random four-fold rotations (increments of $90^\circ$) and randomly cropped to the native resolution of the image encoder, maintaining $\sim 20\%$ of the area of the original image, at each training step.
%
Given the relatively modest size of the fine-tuning dataset, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were observed to be susceptible to overfitting.
%
The temperature hyperparameter $\tau$ was initialized to its pre-trained value.
%
We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

We explore three different methods of training the model on our domain dataset: \emph{(1)} Fine-tuning the entire network, starting from the pre-trained base model; \emph{(2)} Freezing the base image and text encoders, and training a small projection head; and \emph{(3)} Training the entire model from scratch.
%
For \emph{(2)}, we use a 2-layer MLP with 1024 hidden units and a GELU activation layer, projecting onto the 512-dimensional embedding space.

All models were trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate $10^{-5}$ and weight decay $10^{-3}$.
%
Training takes approximately 3 hours on 4 Nvidia A100 GPUs.
Models were instantiated using \package{Transformers} \citep{wolf2019huggingface} and trained using packages from the \package{Jax} \citep{jax2018github} ecosystem.
%

\subsection{Evaluation Metrics}
\label{sec:eval}
